### 2018-09-13 doc/methods.tex
filling in more of results with supplement stuff

This logistic regression model, minus the final selection of priors, is expressed as 
\begin{equation}
  \begin{aligned}
    t_{i} &\sim \text{Bernoulli}(\Theta) \\
    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + a_{d[i], p[i]}) \\
    B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
    \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
    a_{w, p} &\sim \mathrm{N}(\delta_{p}, \sigma_{a}) \\
    \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
  \end{aligned}
  \label{eq:core}
\end{equation}
with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.

@@

In total, these prior choices can be expressed as
\begin{equation}
  \begin{aligned}
    \mu_{d} &\sim 
      \begin{cases}
        N(0, 10) & \text{if } d = 1 \\
        N(-1, 3) & \text{if } d = 2 \\
        N(0, 3) & \text{if } d > 2 \\
      \end{cases}
    \delta &\sim N(0, 1) \\
    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
    \tau_{B} &\sim C^{+}(5) \\
    \tau_{\alpha} &\sim C^{+}(5) \\
    \Omega_{B} &\sim LKJ(1) \\
    \Omega_{\alpha} &\sim LKJ(1) \\
  \end{aligned}
  \label{eq:priors}
\end{equation}

@@

One of the issues with our data set is extreme class imbalance; this means that there are many more survival observations than extinction observations. When the classes are not approximately balanced, the base probability of 1 versus 0 is below 0.5. When the intercept term of the logistic regression is sufficiently negative (-2 or less), the only way to obtain a linear predictor value of 0.5 or greater is if the covariate effects are large or there the covariates are at extreme values. The 0.5 default cut-off point for binary prediction is much to high when events are rare. A method for determining the new ``optimal'' cut-point involves calculating the receiver operating characterist (ROC) curve from the linear predictor estimates and then identifying the point on the ROC curve which maximizes both true positive rate and the false positive rate. The new cut-point is then back-calculated from the true positive and false positive rates. Class membership is then predicted given this new cut-point. This process is done for every posterior draw, as each realization from the posterior predictive distribution has its own optimal cut-point.

>>>>

This logistic regression model, minus the final selection of priors, is expressed as 
\begin{equation}
  \begin{aligned}
    t_{i} &\sim \text{Bernoulli}(\Theta) \\
    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{d[i], p[i]}) \\
    B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
    \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
    % double check this\dots is A MVN dist?
    A_{w, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
    \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
  \end{aligned}
  \label{eq:core}
\end{equation}
with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.

@@

In total, these prior choices can be expressed as
\begin{equation}
  \begin{aligned}
    \mu_{d} &\sim 
      \begin{cases}
        N(0, 10) & \text{if } d = 1 \\
        N(-1, 3) & \text{if } d = 2 \\
        N(0, 3) & \text{if } d > 2 \\
      \end{cases}
    \delta &\sim N(0, 1) \\
    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
    \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
    \tau_{B} &\sim C^{+}(5) \\
    \tau_{\alpha} &\sim C^{+}(5) \\
    \tau_{A} &\sim C^{+}(5) \\
    \Omega_{B} &\sim LKJ(1) \\
    \Omega_{\alpha} &\sim LKJ(1) \\
    \Omega_{A} &\sim LKJ(1) \\
  \end{aligned}
  \label{eq:priors}
\end{equation}

@@

One of the issues with our data set is extreme class imbalance; this means that there are many more survival observations than extinction observations. When the classes are not approximately balanced, the base probability of 1 versus 0 is below 0.5. When the intercept term of the logistic regression is sufficiently negative (-2 or less), the only way to obtain a linear predictor value of 0.5 or greater is if the covariate effects are large or there the covariates are at extreme values. The 0.5 default cutpoint for binary prediction is much to high when events are rare. A method for determining the new ``optimal'' cutpoint involves calculating the receiver operating characteristic (ROC) curve from the linear predictor estimates and then identifying the point on the ROC curve which maximizes both true positive rate and the false positive rate. The new cutpoint is then back-calculated from the true positive and false positive rates. Class membership is then predicted given this new cutpoint. This process is done for every posterior draw, as each realization from the posterior predictive distribution has its own optimal cutpoint.

### 2018-09-13 doc/results.tex
filling in more of results with supplement stuff

@@

% average hazard

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/hazard_baseline}
  \caption{<+caption text+>}
  \label{fig:hazard_baseline}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/hazard_bygroup}
  \caption{<+caption text+>}
  \label{fig:hazard_bygroup}
\end{figure}

@@

There are then two non-nested sources of the second ``level'' of variance: intercept and slopes varying by time of observation, and a mean-zero age at observation varying intercept. Each of the elements of the vector of scale terms \(\tau_{B}\) describe the within-phyla variance for the intercept and the slope terms for the covariates. Similarly, the scale term \(\sigma_{a}\) is a scalar that describes the within-phyla variance of the effect of age at observation.

>>>>

As described above, our data is extremely unbalanced with respect to the number of extinction events relative to the total number of observations; this reality means that 0.5 is an inappropriate cutpoint for estimating if a species observation is their last (i.e. goes extinct). New, optimal cutpoint were calculated from the ROC curve as the level which maximized the true and false positive rates.
The distributions of optimal cut-off point estimates for each of the four models are broadly similar, with all 50\% credible intervals overlapping.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/cut_plot}
  \caption{Posterior estimates of the optimal cutpoints for each of the three models. The optimal cutpoint corresponds to the point which maximizes the true and false positive rates.}
  \label{fig:cut_point}
\end{figure}

@@

% average hazard
The hazard function describes how age of observation impacts probability of extinction. Hazard is estimated for each phylum for the range of ages observed within the phylum. The matrix \(A\) describes the effect of age on extinction probability for age at observation (rows) by phylum (columns). The phylum-level averages of the effect of age of observation on extinction \(\delta\) are themselves drawn from a distribution centered around 0.

Figure \ref{fig:hazard_baselin} is a graph of the average hazard associated with age of observation. This graph illustrates how average extinction risk increases with age for the first few million years before leveling off. A species reaches highest age-related probability of extinction at 10 million years, after which the median becomes volatile until a duration over 30 million years. The flat-ness of the median estimate and the homogeneity of variance durations between approximately 30 million years to under 60 million years is due to the lack of species having survived over 30 million years, which inherently means that these estimates are based on less information and are shrunk towards the overall average while reflecting more and more of our prior uncertainty over time.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/hazard_baseline}
  \caption{Average hazard estimates based on all all observed species ages in millions of years. Hazard in discrete-survival analysis is the probability of going extinct at a given age having survived up to that age. Overall hazard was calculated using the phylum average intercept and assuming all covariates equaled 0. The red line is the median hazard estimate along with 50\%, 80\%, and 95\% credible intervals.}
  \label{fig:hazard_baseline}
\end{figure}

The hazard associated with the individual phyla have broad similarities with the overall hazard estimates (Fig. \ref{fig:hazard_bygroup}). In all four cases, hazard increases over the first 10 million years of a species duration. However, the rate of increase in hazard varies between the phyla, with forams and radiolarians having the most pronounced increase with age. Radiolarians also have the lowest initial hazard of the phyla, followed by forams and calcareous nannoplankton. In contrast, diatoms have the highest initial hazard of the phyla while having the smallest change hazard with age.

Optimal cutpoint between 0.1 and 0.11 percent. We estimate that the effect of age of observation on extinction can reach this threshold after XX My at ##\% posterior probability.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/hazard_bygroup}
  \caption{Hazard estimates for each of the phyla analyzed here. The red line is the median hazard estimate along with 50\%, 80\%, and 95\% credible intervals.}
  \label{fig:hazard_bygroup}
\end{figure}

@@

There are then two non-nested sources of the second ``level'' of variance: intercept and slopes varying by time of observation, and a mean-zero age at observation varying intercept. Each of the elements of the vector of scale terms \(\tau_{B}\) describe the within-phyla variance for the intercept and the slope terms for the covariates. Similarly, the scale term \(\sigma_{A}\) is a scalar that describes the within-phyla variance of the effect of age at observation.
